---
title: "Bicycle"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```



## Create a presentation

* Render this document with "Knit" -- as usual.
* Change the output to `ioslides_presentation` and knit again.



## Packages

```{r}
library(tidyverse)
library(here)
library(fs)
library(vroom)
```



## Definition of tidy data

<img src=http://i.imgur.com/SwZLZ3I.png width=750>



## From messy to tidy

<img src=http://i.imgur.com/ybegKuU.png width=750>



## Import

Import all datasets in "data/by-continent" into a single data frame: `messy`.

```{r message=FALSE}
paths <- dir_ls(here("data", "by-continent"))
messy <- vroom(paths)
messy
```



## Tidy with `pivot_longer()`: Task

* Use `where(is.numeric)` inside `pivot_longer()` to pivot over numeric columns.
* Use the argument `names_to"` to create the new column "metric".

```{r longer-1, eval=FALSE}
longer <- messy %>% 
  ____________(where(__________), ________ = "metric")

longer
```



## Tidy with `pivot_longer()`: Result

```{r longer-2, echo=FALSE}
longer <- messy %>% 
  pivot_longer(where(is.numeric), names_to = "metric")

longer
```



## `separate()`: Task

* `metric` is still messy; `separate()` `metric` into `c("metric", "year")`.

```{r separate-1, eval=FALSE}
tidy <- longer %>% ________(______, c("metric", "____"))
tidy
```



## `separate()`: Result

```{r separate-2, echo=FALSE}
tidy <- longer %>% separate(metric, c("metric", "year"))
tidy
```



## `unite()` and `pivot_wider()`: Task

Revert what you did before: go back to a messy dataset:

* Use `unite()` to unite the columns `metric` and `year` as "metric".
* Use `pivot_wider()` to take the `names_from` the column `metric` and create new columns, taking `values_from` the column `value`.

```{r mess-up-1, eval=FALSE}
tidy %>% 
  _____("metric", ______, year) %>% 
  ___________(names_from = ______, values_from = _____)
```



## `unite()` and `pivot_wider()`: Result

```{r mess-up-2, echo=FALSE}
tidy %>% 
  unite("metric", metric, year) %>% 
  pivot_wider(names_from = metric, values_from = value)
```



## Create a small dataset: Review

Let's create a small dataset to play with. You should understand this code:

- Remove `continent`.
- Subset life expectancy values for Argentina and Germany before 1962.
- Widen the dataset adding new columns with `names_from` the column `metric`.



## Create a small dataset: Review

```{r}
subset1 <- tidy %>% 
  select(-continent) %>% 
  filter(
    country %in% c("Argentina", "Germany"), 
    metric == "lifeExp",
    year < 1962
  ) %>% 
  pivot_wider(names_from = metric)

subset1
```



## Create a small dataset: Task

Now add a new column `mean`, holding the mean `lifeExp` for each `country`:

* Use `group_by()`.
* Use `mutate()` to calculate mean `lifeExp` and to make `year` numeric.
* `ungroup()`.

```{r group-mutate-1, eval=FALSE}
subset2 <- subset1 %>% 
  ________(country) %>% 
  ______(mean = ____(lifeExp), year = as.numeric(____)) %>% 
  _______()

subset2
```



## Create a small dataset: Result

```{r group-mutate-2, echo=FALSE}
subset2 <- subset1 %>% 
  group_by(country) %>% 
  mutate(mean = mean(lifeExp), year = as.numeric(year)) %>% 
  ungroup()

subset2
```



## Create a small dataset: Tweak

Let's degrade this dataset for a later example.

```{r}
arg_ger <- subset2 %>% slice(-2) %>% select(-mean)
```



## `complete()`: What's missing?

Say you have the dataset `arg_ger`; it is missing data in between 1952 and 1957, but you know the historical mean for each country. Let's fill the missing data.

```{r}
arg_ger

# Historical mean (1952-2007) of life expectancy
mean_argentina <- 63.5
mean_germany <- 68.3
```



## `complete()`: Task

* Make the implicit missing data explicit with `complete()`.

```{r complete-1, eval=FALSE}
arg_ger %>% ________(_______, year)
```



## `complete()`: Result

```{r complete-2, echo=FALSE}
arg_ger %>% complete(country, year)
```



## `fill`: Task

* Now fill the missing values of `lifeExp` with the value in `mean_argentina`.

```{r fill-1, eval=FALSE}
filled <- arg_ger %>% 
  complete(country, year, fill = list(lifeExp = ______________))

filled
```



## `fill`: Result

```{r fill-2, echo=FALSE}
filled <- arg_ger %>% 
  complete(country, year, fill = list(lifeExp = mean_argentina))

filled
```



## `full_seq()`: Demo

Compare:

```{r}
filled %>% pull(year)
filled %>% pull(year) %>% full_seq(period = 1)
```



## `full_seq()`: Task

* Use `full_seq()` inside `complete()` to complete the full `year` sequence.

```{r all-years-1, eval=FALSE}
full_mean <- filled %>% 
  complete(year = ________(year, period = 1), country = _______)

full_mean
```



## `full_seq()`: Result

```{r all-years-2, echo=FALSE}
full_mean <- filled %>% 
  complete(year = full_seq(year, period = 1), country = country)

full_mean
```



## `case_when()`: Task

Fill `lifeExp` with the historical mean values for each country:

* Use `case_when()` inside `mutate()`.
* The 3 possible results are `mean_argentina`, `mean_germany`, or `lifeExp`.



## `case_when()`: Task

```{r full-1, eval=FALSE}
full <- full_mean %>% 
  mutate(
    lifeExp = _________(
      _____(lifeExp) & country == "_________" ~ mean_argentina,
      is.na(_______) & _______ == "Germany"   ~ ____________,
      # Any other case
      TRUE                                    ~ lifeExp
    )
  )

full
```



## `case_when()`: Result

```{r full-2, echo=FALSE}
full <- full_mean %>% 
  mutate(
    lifeExp = case_when(
      is.na(lifeExp) & country == "Argentina" ~ mean_argentina,
      is.na(lifeExp) & country == "Germany"   ~ mean_germany,
      # Any other case
      TRUE                                    ~ lifeExp
    )
  )

full
```



## Plot: Task

* Make a line-plot of `year` versus `lifeExp`.
* Add a "dotted" line intercepting `y` at the historical mean of each country.
* Use `facet_wrap()` to plot each country in a separate panel.

```{r line-1, eval=FALSE}
full %>% 
  ggplot(aes(____, _______)) + 
  _____line() +
  _____hline(__________ = c(mean_argentina, mean_germany), linetype = "______") +
  __________(~country)
```



## Plot: Task

```{r line-2, echo=FALSE}
full %>% 
  ggplot(aes(year, lifeExp)) + 
  geom_line() +
  geom_hline(yintercept = c(mean_argentina, mean_germany), linetype = "dotted") +
  facet_wrap(~country)
```



# Takeaways


## Takeaways: Imoprt

* Use `vroom()` to ready multiple files at once and into a single data frame.



## Takeaways: Tidy

* The tidyverse prefers tidy data where:
  * Every variable is in its own column.
  * Every observation is in its own row.
* Use `pivot_longer()` and `pivot_wider()` to move columns to values and back.
* Use `unite()` to unite multiple ones into a single one.
* Use `separate()` to separate one column into multiple ones.
* Complete missing data with `complete()`, `full_seq()`, `fill`, `case_when()`.



## Takeaways: Transform

* The select-helpers (e.g. `where()`) appear in many places in the tidyverse.
* You can `filter()` with multiple conditions separated by `&` or comma `,`.



## Takeaways: Visualise

* You can extend the basic plot template to include multiple layers.
* Use `facet_wrap()` to plot subsets of data in separate panels.



## Takeaways: Communicate

* You can change the output format to, e.g., create .html presentations.
